\newpage
\addchap{List of publications and/or patents and conference participation}
\label{chap:publications}
\addsec*{(1) List of publications and/or patents produced as part of the thesis project:}
\begin{enumerate}
\item \textbf{Düll, R.}, Bufferand, H., Serre, E., Ciraolo, G., Quadri, V., Rivals, N. and Tamain, P., 2024. Introducing electromagnetic effects in Soledge3X. \textit{Contributions to Plasma Physics}, p.e202300147.
\item \textbf{Düll, R.}, Ciraolo, G. Bufferand, H., Serre, E.,, Quadri, V., Rivals, N., Tamain, P., Sureshkumar, S. and Varadarajan, N. Implementation of a non-axisymmetric magnetic configuration in SOLEDGE3X to simulate 3D toroidal magnetic ripple effects: Application to WEST. \textit{Nuclear Materials and Energy} [under review]
\item \textbf{Düll, R.}, Bufferand, H., Serre, E., Ciraolo, G., Quadri, V., Rivals, N., Schwander, F. and Tamain, P., An electromagnetic model in SOLEDGE3X for edge plasma turbulence simulations in tokamak. \textit{Journal of Computational Physics} [to be submitted]
\item Quadri, V., Tamain, P., Marandet, Y., Bufferand, H., Rivals, N., Ciraolo, G., Falchetto, G., \textbf{Düll, R.} and Yang, H., 2024. Self‐organization of plasma edge turbulence in interaction with recycling neutrals. \textit{Contributions to Plasma Physics}, p.e202300146.
\item  Quadri, V., Tamain, P., Marandet, Y., Bufferand, H., Rivals, N., Ciraolo, G., Falchetto, G.L., \textbf{Düll, R.}, Sureshkumar, S., Varadarajan, N. and Yang, H., 2024. Edge plasma turbulence simulations in detached regimes with the SOLEDGE3X code. \textit{Nuclear Materials and Energy}, p.101756.
\item De Gianni, L., Ciraolo, G., Giruzzi, G., Falchetto, G., Rivals, N., Balbinot, L., Varadarajan, N., Sureshkumar, S., Artaud, J.F., Bufferand, H. and \textbf{Düll, R.}, Core and edge modeling of JT-60SA H-mode highly radiative scenarios using SOLEDGE3X-EIRENE and METIS codes. \textit{Frontiers in Physics}, 12, p.1422286.
\end{enumerate}

\newpage

\addsec*{(2) Participation in conferences and summer schools during the thesis period:}
\begin{enumerate}
\item Summer School PlasmaSurf by Instituto de Plasmas e Fus\~{a}o Nuclear. July 2022, Lisbon, Portugal
\item Conference EPS - 49th European Conference on Plasma Physics, European Physical Society. July 2023, Bordeaux, France. Poster presenation: \textit{"A new electromagnetic model in SOLEDGE3X"}
\item Conference PET19 - 19th International Workshop on Plasma Edge Theory in Fusion Devices. September 2023, Hefei, China. Contributed talk: \textit{"Introducing electromagnetic effects in Soledge3X"}
\item Conference PSI-26 - 26th International Conference on Plasma Surface Interaction in Controlled Fusion Devices. May 2024, Marseille, France. Poster presentation: \textit{"Electromagnetic effects on turbulent structures in edge plasmas with SOLEDGE3X"}
\item EUROfusion - TSVV3 Annual Workshop. May 2024, Leuven, Belgium [remote participation]. Presentation: \textit{"Simulating electromagnetic effects in SOLEDGE3X"}
\item Conference ECCOMAS - 9th European Congress on Computational Methods in Applied Sciences and Engineering. June 2024, Lisbon, Portugal. Participation at the mini-symposium: Magnetohydrodinamic Numerical Modeling of Magnetised Plasmas. I with the oral presentation: \textit{"An electromagnetic model for SOLEDGE3X"}
\item EUROfusion - TSVV1 Progress Workshop. September 2024, Garching, Germany. Presentation: \textit{"A new electromagnetic model for turbulent simulations in SOLEDGE3X"}
\end{enumerate}