\addchap{Acknowledgments}

I want to thank IRFM and Aix-Marseille University for allowing me to pursue and complete this PhD thesis over the past three years. It has been a journey with ups and downs, stretching across moments of fervent excitement and nerve-wracking struggles.

I am very grateful to my academic supervisor, Eric Serre, for his guidance and support throughout this process. His encouragement, foresight, and feedback have been invaluable for this work. He knew what to prioritize and focus on at any moment, and as such I could stay on course with a clear goal before me.

Equally, my advisor, Hugo Bufferand, was my daily support throughout this endeavour. He accompanied me from my first steps in the world of plasma physics to deep and precise questioning. I truly enjoyed our many conversations about the requirements and possibilities of the model and its best implementation. Your clarity in explaining complex concepts and availability to address my endless questions have been priceless. This thesis owes much to your guidance and support.

I want to thank Patrick Tamain for his help with the code framework and all the caveats I had to be aware of. I am equally grateful for his perspective and feedback on the physical model I was working on. I never forget our endless but fruitful Monday morning developer discussions. Without your involvement, my developments would undoubtedly have remained buried in a forgotten branch of the repository and unseen by any potential user. 

I am also grateful for the enriching discussions on plasma theory with various institute colleagues. First, I would like to thank Frédéric Schwander, whose insights into fluid models helped clarify many aspects of my work. Yanick Sarazin, I appreciated our conversations about the theory of edge plasma turbulence dynamics. Finally, Eric Nardon, you helped to bridge the gap between the new electromagnetic fluid approach and state-of-the-art MHD models.

I want to thank the members of my PhD jury for taking the time to read my manuscript, traveling to Cadarache for those who could, and asking thoughtful questions during the defense. 

During my time at CEA, I had the chance to meet some fantastic people who made this journey very enjoyable. We almost lived like a small family in the unpretentious building 513, safeguarded by our ever-watchful, hairy boar guardians. Virginia, Yann, Olivier, Mattia, Jai, Kévin, Robin, Nicolas, Hao, Emily, Srikanth, Ludovica, Roméo, Naren - we all know what it means to lose the mind on a single question for several years. And yet, we still found our moments of fun, whether during the coffee breaks or over beers in the uncountable Plasmapéros. For my two successive office mates, Robin and Srikanth, I apologize for driving you crazy with my constant random remarks; Virginia, thanks for being the best travel mate on the epic journey to China; Ludovica, you taught me more Italian than anything I learned during the PhD. I even talked to people in the neighbouring buildings! Quentin, with whom I ended up cycling along the Mediterranean coast; Théo, we still need to beat skiing records; Mathieu, soon will come the day we open a restaurant. A special note to Michele, who just arrived in this parallel universe: best of luck—I’m sure you’ll do great! I also want to acknowledge the people at the M2P2 lab. I have great memories of the conference trip to Portugal with Eduardo, Hippolyte, and Tom. Although I didn’t visit the lab as often as I’d have liked, I’m looking forward to reconnecting more during my postdoc.

Moving to Marseille for my PhD turned out to be one of the best decisions I ever made. I quickly grew to love the city, its vibrant atmosphere, and the wonderful people I met there. I’ll always have fond memories of my time there.

Finally, a big thank you to my family for always being there for me. Your support has meant a lot.

And last but not least, thanks to you, the reader, for making it this far. The fact that you took the time to read this thesis really means a lot to me, and I know you will read past this Acknowledgements section. Thank you!